---
title: "EMTscoreData Package"
author: "haimei wen"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{EMTscoreData Vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# 1. Introduction of EMTscoreData
EMTscoreData is an R ExperimentHub package that provides curated gene expression datasets for benchmarking and developing epithelial–mesenchymal transition (EMT) scoring methods. In future updates, we will add more EMT datasets to facilitate users in conducting EMT-related research. This package is now prepared as the companion dataset for the upcoming EMTscore R package.

# 2. Installation
## Install the development version from GitHub
```{r install_github, eval=FALSE}
# install.packages("devtools")
#devtools::install_github("wenmm/EMTscore")
#devtools::install_github("wenmm/EMTscoreData")
```

## Install from Bioconductor (recommended once available):
```{r install}
#if (!require("BiocManager", quietly = TRUE))
#    install.packages("BiocManager")

#BiocManager::install("EMTscore")
#BiocManager::install("EMTscoreData")
```

# 3. Download data from the EMTscoreData R package
You can obtain the datasets included in this package through the `ExperimentHub` framework. To start, use query() to inspect all records associated with the `EMTscoreData` package, which will show you the available resources along with their corresponding identifiers. After identifying the items of interest, you can download them directly from the hub.
```{r}
library(EMTscoreData)
#library(EMTscore)
library(ExperimentHub)
library(SummarizedExperiment)
library(Seurat)
library(ggplot2)
```

## 3.1 Browse all EMTscoreData records
EMTscoreData registers multiple resources (datasets) on the hub; you can list them first, then download specific ones by their unique ExperimentHub IDs (e.g., EH10291).

`query()` returns a Hub object listing all records matching the keyword.
This helps you discover what datasets are available and what their IDs are.
```{r}
eh <- ExperimentHub()
query(eh , 'EMTscoreData')
```

## 3.2 Download data using the unique identifier:
Below we download three example datasets. These objects are typically SingleCellExperiment, which is the recommended Bioconductor container for single-cell expression data.
```{r}
A549_TNF <- eh[['EH10291']]
A549_EGF <- eh[['EH10292']]
A549_TGFB1 <- eh[['EH10293']]
```

Sanity checks: show the class and a brief overview
```{r}
class(A549_TNF)
A549_TNF
```
## 3.3 What is inside a SingleCellExperiment?
A SingleCellExperiment generally contains:
assays: expression matrices (e.g., counts, logcounts)
rowData: feature-level annotation (genes)
colData: cell-level metadata (e.g., pseudotime, condition, batch)
(optional) reducedDims: PCA/UMAP embeddings, etc.
Let’s inspect what assays and metadata columns exist in one dataset:
```{r}
#List available assays (common ones: "counts", "logcounts")
assayNames(A549_TNF)
```

```{r}
#List available per-cell metadata fields
colnames(colData(A549_TNF))
```

If your dataset includes a pseudotime field, it will typically appear in colData(sce)$Pseudotime (or a similarly named column). The plotting example later assumes a pseudotime column exists; if your column name differs, adjust the time_col argument accordingly.

# 4. Data analysis and visualization
Some single-cell datasets can be large. Running full analyses on all cells in a vignette may exceed RAM on a typical machine and may fail during package checks.
Therefore, this vignette uses a small subset of cells from each dataset. This keeps computations fast and demonstrates the workflow without requiring high-memory hardware.
```{r}
set.seed(1)

subset_cells <- function(sce, n = 500) {
n <- min(n, ncol(sce))
sce[, sample(seq_len(ncol(sce)), n)]
}

A549_TNF_small <- subset_cells(A549_TNF, n = 1500)
A549_EGF_small <- subset_cells(A549_EGF, n = 1500)
A549_TGFB1_small <- subset_cells(A549_TGFB1, n = 1500)

objects <- list(
A549_TGFB1 = A549_TGFB1_small,
A549_EGF = A549_EGF_small,
A549_TNF = A549_TNF_small
)

#Confirm the reduced size

sapply(objects, ncol)
```
## 4.1 Load an EMT gene set (GMT) from Zenodo
Many EMT scoring methods require a gene set. Here we demonstrate how to download a GMT file from Zenodo and parse it into a simple character vector of genes.

Download the GMT file:
```{r}
library(BiocFileCache)
url <- "https://zenodo.org/records/18168504/files/HALLMARK_EPITHELIAL_MESENCHYMAL_TRANSITION.v2025.1.Hs.gmt"

bfc <- BiocFileCache() # creates/uses a cache folder
gmt_file <- bfcrpath(bfc, url) # downloads once; returns the cached file path
```

Parse GMT and extract EMT genes:
A GMT file is tab-delimited:
`GeneSetName \t Description \t GENE1 \t GENE2 \t ...`
The helper below extracts all genes across gene sets in the GMT and returns unique gene symbols.
```{r}
read_gmt <- function(fname){
  #Read all lines from the GMT file
  gmt_lines <- readLines(fname)
  #Split each line by tab; each line becomes a character vector
  gmt_list <- lapply(gmt_lines, function(x) unlist(strsplit(x, split="\t")))
  #For each line, genes start at element 3 (after name and description)
  gmt_genes <- lapply(gmt_list, function(x) x[3:length(x)])
  #Get unique gene list
  genes <- unique(unlist(gmt_genes))
  #Return a data frame whose column name is `gene` and whose values are gene names.
  return(data.frame(gene = genes, stringsAsFactors = FALSE))
}

Genesets <- read_gmt(gmt_file)
Genesets
```

## 4.2 Calculating EMT Scores for Datasets Stored in EMTscoreData
Calculated EMT scores can then be plotted against pseudotime to examine dynamic changes in epithelial–mesenchymal states.

Because this package is a data package, we keep the computation simple and robust for vignettes.
Here we compute a demo EMT score defined as the mean log-expression across EMT genes for each cell:

Input: list of SingleCellExperiment/Seurat object + EMT gene vector
Output: a list of Seurat objects, each containing a new metadata column with one score per cell.
```{r}
## ------------------------------------------------------------------
## Prepare input objects
## ------------------------------------------------------------------
## We first organize the datasets into a named list. Each element
## corresponds to one experimental condition and contains a Seurat
## object (or a SingleCellExperiment object that will be converted
## internally). This unified structure allows us to apply the same
## EMT scoring procedure to multiple datasets in a consistent way.

objects <- list(
A549_TGFB1 = A549_TGFB1_small,
A549_EGF   = A549_EGF_small,
A549_TNF   = A549_TNF_small
)

## EMT gene list extracted from the GMT file. These genes define
## the epithelial–mesenchymal transition (EMT) signature used
## for score calculation.
emt_genes <- Genesets$gene


## ------------------------------------------------------------------
## Function: add_EMT_score
## ------------------------------------------------------------------
## This function computes an EMT score for each cell in a collection
## of Seurat (or SingleCellExperiment) objects using a predefined
## EMT gene set.
##
## Input:
##   - objects: a named list of Seurat objects or SingleCellExperiment
##              objects. Each object represents one dataset or condition.
##   - gmt_file: path to a GMT file containing EMT-related genes.
##   - emt_name: name of the metadata column that will store the EMT score.
##
## Output:
##   - A list of Seurat objects, each containing a new metadata column
##     with one EMT score per cell.
##
## Note:
##   For demonstration purposes in this vignette, the EMT score is
##   calculated using Seurat's AddModuleScore function. More advanced
##   or alternative EMT scoring methods are implemented in the
##   companion EMTscore package.

add_EMT_score <- function(objects,
                          gmt_file,
                          emt_name = "EMT_Score") {
  
  ## Read the EMT gene set from the GMT file
  Genesets <- read_gmt(gmt_file)
  emt_genes <- Genesets$gene
  
  ## Apply EMT scoring to each object in the list
  obj_list <- lapply(names(objects), function(name) {
    
    obj <- objects[[name]]
    
    # Convert SCE → Seurat
    if (inherits(obj, "SingleCellExperiment")) {
      message("Converting SCE to Seurat: ", name)
      obj <- as.Seurat(obj, data = "logcounts")
    }
    if (!inherits(obj, "Seurat")) {
      stop("Object ", name, " is not a Seurat or SCE object.")
    }
    
    ## Update the Seurat object to the current structure if needed.
    obj <- UpdateSeuratObject(obj)
    
    # get gene expression matrix
    geneExp <- GetAssayData(obj, assay = "RNA", layer = "data")
    
    ## Compute the EMT score using the EMT gene set.
    ## AddModuleScore returns a column named '<emt_name>1', which we
    ## subsequently rename for clarity.
    obj <- AddModuleScore(obj, features = list(emt_genes), name = emt_name, ctrl = 5)
    old <- paste0(emt_name, "1")
    obj <- SeuratObject::AddMetaData(
      object   = obj,
      metadata = obj[[old]][, 1, drop = TRUE],  # 提取为向量
      col.name = emt_name
    )
    obj[[old]] <- NULL
    return(obj)
  })
  
  
  ## Preserve original dataset names
  names(obj_list) <- names(objects)
  return(obj_list)
}


## ------------------------------------------------------------------
## Function: plot_EMT_from_objects
## ------------------------------------------------------------------
## This function visualizes EMT scores as a function of pseudotime
## across multiple datasets.
##
## Input:
##   - obj_list: a list of Seurat objects produced by add_EMT_score().
##   - col_name: name of the metadata column representing pseudotime.
##   - emt_score_col: name of the metadata column storing EMT scores.
##
## Output:
##   - A ggplot object showing smoothed EMT score trends along pseudotime
##     for each experimental condition.

plot_EMT_from_objects <- function(obj_list, col_name,
                                  emt_score_col) {
  
  ## Merge metadata from all Seurat objects into a single data frame
  ## for visualization.
  plot_df <- do.call(rbind, lapply(names(obj_list), function(name) {
    obj <- obj_list[[name]]
    df <- obj[[]][, c(col_name, emt_score_col), drop = FALSE]
    colnames(df) <- c(col_name, emt_score_col)
    ## Track the experimental condition
    df$Condition <- name
    ## Order cells by pseudotime for smoother visualization
    df <- df[order(df[[col_name]]), ]  
    df
  }))
  
  # Draw smooth curves
  p <- ggplot(plot_df, aes_string(x = col_name, y = emt_score_col, 
                                  color = "Condition")) +
    geom_smooth(method = "loess", se = FALSE, linewidth = 1.2) +
    theme_classic(base_size = 14) +
    labs(x = col_name, y = emt_score_col, color = "Condition")
  
  return(p)
}

## ------------------------------------------------------------------
## Run EMT scoring and visualization
## ------------------------------------------------------------------
## Compute EMT scores for all datasets and visualize their dynamics
## along pseudotime.

seurat_objs <- add_EMT_score(objects, 
                             gmt_file = gmt_file,
                             emt_name = "EMT_score")
p_Seurat <- plot_EMT_from_objects(seurat_objs, 
                                  col_name = "Pseudotime", 
                                  emt_score_col = "EMT_score")
p_Seurat
```


Session information:
```{r}
sessionInfo()
```
